Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 paper.tex
Update on SAIGE implementation
Closes #173

\subsection{Interoperability: vcztools}
To provide compatibility with existing workflows we have
developed the \texttt{vcztools} command line utility which
implements a subset of \texttt{bcftools} functionality.
Vcztools is intended to be a drop-in replacement for \texttt{bcftools},
providing an evolutionary path for VCF Zarr adoption,
with compatibility being a core objective.
Vcztools currently supports Zarr stored in the local file system
and Zip archives, and support for major cloud stores will be added in the near
%TODO Resolve this note
future. [JK: note this is a pretty simple feature, maybe we should just try
to add it?]
Thus, users can convert their VCF data to Zarr using \texttt{bio2zarr},
store it in a local file system, Zip file or cloud store as desired
and then access this data source within existing processing pipelines,
essentially replacing \texttt{bcftools view <vcf file>} with
\texttt{vcztools view <Zarr URL>}.
This allows established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

@@

To form a practical replacement for bcftools in production pipelines
it is important that \texttt{vcztools} can produce VCF text at a reasonable
rate. To do this we measure the time taken to generate VCF text for the
last 10,000 variants in the dataset (using the previously extracted
position column to determine the coordinates), \texttt{vcztools view
chr21\_10\_5.vcz -r1:47958263-}). This took 30s, at a rate of 124.3 MiB
of VCF text per second. The bcftools command on the corresponding
BCF file took 25s (152.64 MiB/s) and on the VCF file it took 63s (60.57 MiB/s).

% TODO texttt
Vcztools is primarily intended to used as a replacement for bcftools
in existing workflows based on pipes, where VCF text is sequentially
filtered and processed by different processes in the classical
Unix manner. In this context the most important property is that vcztools
can produce VCF text as quickly as other processes in the pipeline can consume
it. To get an estimate of the rate at which VCF can be parsed, we wrote a simple
program that reads a VCF file and parses it line-by-line using the \texttt{htslib}
API~\citep{bonfield2021htslib}. For the 10,000 variant subset discussed
above (3.73 GiB) this took 31s, giving a parsing rate of 123.1 MiB/s
(almost identical to the write-rate of vcztools above). Given this limiting
factor on overall throughput,
it is unlikely that \texttt{vcztools view} would be a bottleneck
in practical processing pipelines.

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing
high-performance analysis tools,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. 
Using the TensorStore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend 
(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats) 
for which the core functionality is defined in a 264 line C++ file.

To assess the performance of this implementation we compared the time required
to perform an association test using the simulated dataset of 100,000 samples
from Fig~\ref{fig-data-storage}. We performed a single-variant association
test based on a phenotype simulated using
\texttt{tstrait}~\citep{tagami2024tstrait},
and measure the overall time required when using the 
equivalent VCF, BCF, SAV and Zarr files.

We compared the same analysis using the existing VCF, BCF and Savvy backends
to our prototype Zarr backend.
[Times: BCF 70m, VCF 162m, Savvy 7m, Zarr 20m]

>>>>

\subsection{Interoperability: vcztools}
To provide compatibility with existing workflows we have
developed the \texttt{vcztools} command line utility which
implements a subset of \texttt{bcftools} functionality.
Vcztools is intended to be a drop-in replacement for \texttt{bcftools},
providing an evolutionary path for VCF Zarr adoption,
with compatibility being a core objective.
Vcztools currently supports Zarr stored in the local file system
and Zip archives, and support for major cloud stores will be added in the near future. 
Thus, users can convert their VCF data to Zarr using \texttt{bio2zarr},
store it in a local file system, Zip file or cloud store as desired
and then access this data source within existing processing pipelines,
essentially replacing \texttt{bcftools view <vcf file>} with
\texttt{vcztools view <Zarr URL>}.
This allows established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

@@

To form a practical replacement for bcftools in production pipelines
it is important that \texttt{vcztools} can produce VCF text at a reasonable
rate. To do this we measure the time taken to generate VCF text for the
last 10,000 variants in the dataset (using the previously extracted
position column to determine the coordinates), 
\texttt{vcztools view
chr21\_10\_5.vcz -r1:47958263-}. This took 30s, at a rate of 124.3 MiB
of VCF text per second. The bcftools command on the corresponding
BCF file took 25s (152.64 MiB/s) and on the VCF file it took 63s (60.57 MiB/s).

% TODO texttt
Vcztools is primarily intended to used as a replacement for bcftools
in existing workflows based on pipes, where VCF text is sequentially
filtered and processed in the classical
Unix manner. In this context the most important property is that vcztools
can produce VCF text as quickly as other processes in the pipeline can consume
it. To get an estimate of the rate at which VCF can be parsed, we wrote a simple
C program that reads a VCF file and parses it line-by-line using 
\texttt{htslib}~\citep{bonfield2021htslib}. 
For the 10,000 variant subset discussed
above (3.73 GiB) this took 31s, giving a parsing rate of 123.1 MiB/s
(almost identical to the write-rate of vcztools above). Given this limiting
factor on overall throughput,
it is unlikely that \texttt{vcztools view} would be a bottleneck
in practical processing pipelines.

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing
high-performance analysis tools,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. 
Using the TensorStore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend 
(SAIGE reads data from PLINK, BGEN, VCF, BCF and SAV formats),
for which the core functionality is defined in a 264 line C++ file.
TensorStore provides efficient read and write access to array
data in several formats, used, for example, in
checkpointing~\cite{orbax-tensorstore} 
TensorFlow model training~\citep{tensorflow2015-whitepaper}.
It supports several cloud platforms, and is based on an
asyncronous, parallel processing model (see Discussion).

To assess the performance of this implementation we compared the time required
to perform an association test using the simulated dataset of 100,000 samples
from Fig~\ref{fig-data-storage}. 
We performed a single-variant association
test based on a phenotype simulated using
\texttt{tstrait}~\citep{tagami2024tstrait},
and measured the overall time required when using the 
equivalent VCF (162m), BCF (70m), SAV (7m) and Zarr (20m) files.
Thus, in this case, substantially improved performance over the 
most popular formats, along with native cloud-store support are 
possible with relatively minor coding effort.
We validated the implementation by comparing the association
test results with the output of the other backends,
confirming that they were numerically identical. 
As TensorStore does not currently support string data for Zarr, 
however,
we were not able to fully replicate the output of fields 
such as MarkerID, Allele1 etc.
